---
title: "RBANCOVA.RMST: RBANCOVA for RMST estimates with Categorized time-to-event data"
output:
  github_document: default # put github_document *before* html_document
  html_document: default
vignette: >
  %\VignetteIndexEntry{RBANCOVA.RMST: RBANCOVA for RMST estimates with Categorized time-to-event data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
set.seed(2677)
```



```{r setup, include = FALSE}
# During development, you can uncomment the next line in an interactive session
# to use the functions in R/ without installing the package:
# devtools::load_all("..")
devtools::install()

library(RBANCOVA.RMST)
library(dplyr)
library(ggplot2)
library(stringr)
```

# Overview

This vignette shows how to implement Randmoziation-Based Covariance Analysis (RB_ANCOVA) for Restricted Mean Survival Time (RMST) contrasts using RBANCOVA.RMST. We’ll demonstrate via application to estimating the effect of D-Penicillamine for Primary Biliary Cholangitis (PBC).

## APPLICATION: D-PENICILLAMINE FOR PRIMARY BILIARY CHOLANGITIS
Primary biliary cholangitis (PBC) is a chronic autoimmune disease characterized by progressive inflammation and destruction of the small bile ducts within the liver, ultimately leading to cirrhosis (permanent scarring of liver tissues) and liver failure in advanced stages (May Clinic 2024). The progression of PBC is inevitable, but it can be prolonged with treatment. Due to its clinical importance and the need for effective therapeutic options, PBC has been a focus of clinical trials for several decades. A dataset often used to study the natural history and treatment response in PBC patients originates from a Mayo Clinic trial conducted between 1974 and 1984.

The dataset includes 424 patients diagnosed with PBC who were referred to the Mayo Clinic between 1974 and 1984 (Therneau 2024, Therneau & Grambsch 2000). Of these, 312 patients were enrolled in a randomized, placebo-controlled trial of the drug D-penicillamine, while the remaining 112 cases did not participate in the clinical trial but consented to basic measurements and follow-up for survival. The dataset encompasses clinical variables, including age, sex, and a range of laboratory biomarkers such as serum bilirubin, serum albumin, and alkaline phosphatase levels, alongside clinical indicators like the presence of ascites, hepatomegaly, blood vessel malformations (spider angiomata), and histologic stage of disease at entry. Participants' survival outcomes are tracked until death, loss to follow-up (censored), liver transplantation (considered censored at time of transplant), or censoring at the study endpoint in July 1986. The PBC dataset can be accessed by installing the *survival* package in R with dataset *pbc* (Therneau 2024).

```{r, include=TRUE}

#--------------------------------------------------------------------------------#
# Filter PBC Data
#--------------------------------------------------------------------------------#
library(survival)

#Filter to those enrolled in the randomized trial
pbc_filtered <- pbc %>% filter(!is.na(trt))

#Original data: 0/1/2 = censored/transplant/dead
##Edit so that transplant = censored at time of transplant [death = 1, 0 = censored/transplant]
pbc_filtered <- pbc_filtered %>%
  mutate(status_dichot = ifelse(status == 2, 1, 0))

#Create year variable and create arm variable [0=placebo, 1=D-penicillmain]
pbc_filtered <- pbc_filtered %>%
  mutate(time_year = time/365,
         arm = ifelse(trt == 2, 0, trt))

#Current RBANCOVA-RMST methodology assumes no missing data
#impute mean copper level within each group when missing. 
mean_copper_1 <- mean(pbc_filtered[!(is.na(pbc_filtered$copper)) & pbc_filtered$arm==1,]$copper)
mean_copper_0 <-  mean(pbc_filtered[!(is.na(pbc_filtered$copper)) & pbc_filtered$arm==0,]$copper)

pbc_filtered <- pbc_filtered %>%
  mutate(copper_imp = ifelse(is.na(copper), ifelse(arm==1, mean_copper_1, mean_copper_0), copper))

```


This analysis focuses on the 312 trial participants (158 in the D-penicillamine arm, 154 in the placebo arm). We implement randomization-based covariate adjustment for RMST estimation, as described in Section \ref{sec:ch4_methods}, to estimate treatment effects when adjusting for a selection of covariates. 


The methodology described in Section \ref{sec:ch4_rmst_methodology} was applied to the PBC dataset. The Kaplan-Meier curves below illustrate the time to event for participants in the placebo and D-penicillamine groups over a 12.5-year follow-up period. Over the first five years, there is minimal separation in survival probabilites between the two arms, with the D-penicillamine arm showing slightly higher survival probabilities compared to the placebo group. However, starting around five years, the Kaplan-Meier survival curves demonstrate that participants in the placebo arm generally had a higher probability of survival compared to those in the D-penicillamine group until approximately year nine or ten. 

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.width=12, fig.height=6, out.width="100%"}

library(survminer)
fit_trt <- survfit(Surv(time_year, status_dichot) ~ trt,
                   data =pbc_filtered)
# Visualize with survminer
ggsurvplot(fit_trt, 
           data = pbc_filtered, 
           risk.table = "nrisk_cumcensor", 
           censor.shape = NA,                 # Remove plus signs at censoring points
           legend.labs = c("D-penicillamine", "Placebo"),  # Rename the groups
           xlab = "Years",                     # Change x-axis label to 'Years'
           legend.title = "Trial Arm",
           break.time.by = 1)    
```
***Figure 1.*** *Kaplan-Meier (KM) estimate for time-to-death of each treatment group (D-penicillamine or placebo) for patients enrolled in the PBC Mayo Clinic trial. Time represents years from trial registration to death or last date known alive. Risk table shows number at risk (cumulative number censored) at each time point.*


## Single Timepoint

To quantify the observed difference, we calculated the RMST for each group from the start of the trial until year ten using ten yearly intervals. Time was restricted to ten years of follow-up as many individuals at risk after year ten were censored before the end of follow-up. This yielded an RMST of 7.13 years (SE 0.28) for the D-penicillamine group and 7.28 years (SE 0.30) for the placebo group. The unadjusted difference in RMST (D-penicillamine - placebo) was estimated to be -0.15 years with a confidence interval of [-0.95, 0.65] ***Figure 2***. *P-values are not shown but can be requested for contrasts between treatment groups within the rbancova_rmst() function by adding* ***show_p = TRUE***.

```{r}
pbc_filtered_trt <- pbc_filtered[pbc_filtered$arm==1,]
pbc_filtered_ctl <- pbc_filtered[pbc_filtered$arm==0,]

rbancova_rmst(data_treated = pbc_filtered_trt, data_control=pbc_filtered_ctl, time_var="time_year", event_var="status_dichot", covariate_vars=NULL, timepoints=10, num_intervals=10)

```

To further refine the estimated difference in RMST, we employed RB-ANCOVA to adjust for covariates selected via stepwise selection. The covariates retained for adjustment were histologic stage of disease, copper (ug/day), serum bilirubin (mg/dl), and serum albumin (g/dl), reflecting clinically relevant factors that might impact survival outcomes. Baseline balance between the trial arms was similar for most of these covariates (Table \ref{table:pbc_covs}): the average histoloic stage was slightly higher in the placebo group (3.09 vs. 2.97), while serum bilirubin levels were lower in the D-penicillamine group (2.87 vs. 3.65 mg/dl). Copper levels and serum albumin were comparable between groups, indicating minimal baseline differences. Assessing the extent of random imbalance between the two treatment arms, as described in Section \ref{sec:ch4_Step6_wls_main} gave no evidence of unexpected imbalance among the covariates (p-value = 0.23). 

\begin{table}
\caption{Mean (SE) of selected variables by treatment arm}

\centering
\begin{tabular}{lcccc}
\toprule
\textbf{Trial Arm} & \textbf{Histologic Stage} & \textbf{Copper} & \textbf{Serum Bilirubin} & \textbf{Serum Albumin}\\
\midrule
Placebo & 3.09 (0.07) & 97.65 (6.46) & 3.65 (0.43) & 3.52 (0.03)\\
D-penicillamine & 2.97 (0.07) & 97.64 (7.18) & 2.87 (0.29) & 3.52 (0.04)\\
\hiderowcolors
\hline  % Please only put a hline at the end of the table
\end{tabular}\label{table:pbc_covs}
\end{table}



After adjusting for these covariates using RB-ANCOVA, the adjusted RMST difference was estimated to be -0.44 years with a confidence interval of [-1.05, 0.17].  
While the RB-ANCOVA results are consistent with the unadjusted results, the covariate adjustment improved the precision of the estimate. The CI half-width for the unadjusted estimate was 0.80 versus 0.61 for the adjusted estimate. 

```{r}

results_1time <- rbancova_rmst(data_treated = pbc_filtered_trt, data_control=pbc_filtered_ctl, time_var="time_year", event_var="status_dichot", covariate_vars=c("stage","copper_imp", "bili", "albumin"), timepoints=10, num_intervals=10)

```

***Figure 2*** below also shows the results of implementing continuous RMST methodology, unadjusted, as well as adjusted for covariates via Tian et al. (2018)'s ANCOVA-type method and Andersen et al. (2003, 2004)'s pseudo-value method. Results when implementing the continuous methodology are consistent with those found with our methodology. Adjusting for covariates via RB-ANCOVA results in a similar estimate and confidence interval to that of Tian et al. (2018)'s and Andersen et al. (2003, 2004)'s methods, but the RB-ANCOVA adjustment does not require the assumptions of these model-based methods to hold.


```{r, include = FALSE}
library(pseudo)
library(geepack)
library(survRM2)


#Continuous results

#Unadjusted
cont_unadj <- rmst2(pbc_filtered$time_year, pbc_filtered$status_dichot, pbc_filtered$arm, tau=10)

#Tian
cont_tian <- rmst2(pbc_filtered$time_year, pbc_filtered$status_dichot, pbc_filtered$arm, tau=10, covariates = pbc_filtered[,c("stage","copper_imp", "bili", "albumin")])

#Andersen
pseudo <- pseudomean(time = pbc_filtered$time_yea, event = pbc_filtered$status_dichot, tmax = 10)
fit_pv <- geeglm(pseudo ~ as.factor(arm)+as.factor(stage)+copper_imp+bili+albumin, id = seq_len(nrow(pbc_filtered)),
                 family = gaussian, corstr = "independence", data = pbc_filtered)
summary(fit_pv)  # coef on arm = covariate-adjusted ΔRMST(τ)

#Gather results with CIs from all methods

#RBANCOVA


#Continuous Unadjusted
cont_u_est <- cont_unadj$unadjusted.result["RMST (arm=1)-(arm=0)","Est."]
cont_u_lcl <- cont_unadj$unadjusted.result["RMST (arm=1)-(arm=0)","lower .95"]
cont_u_ucl <- cont_unadj$unadjusted.result["RMST (arm=1)-(arm=0)","upper .95"]

#Tian
tian_est <- cont_tian$adjusted.result["RMST (arm=1)-(arm=0)","Est."]
tian_lcl <- cont_tian$adjusted.result["RMST (arm=1)-(arm=0)","lower .95"]
tian_ucl <- cont_tian$adjusted.result["RMST (arm=1)-(arm=0)","upper .95"]

#Anderesen
alpha <- 0.05
z <- qnorm(1 - alpha/2)
andersen_est <- fit_pv$coefficients["as.factor(arm)1"]
gee_se <- sqrt(diag(vcov(fit_pv))["as.factor(arm)1"])
andersen_lcl <- andersen_est - z*gee_se
andersen_ucl <- andersen_est + z*gee_se


#--- helpers ---------------------------------------------------------------
.get_tbl <- function(x) if (!is.null(x$table_pretty)) x$table_pretty else x
.parse_estCI <- function(s) {
  m <- str_match(s, "([-+]?\\d*\\.?\\d+(?:[eE][-+]?\\d+)?)\\s*\\(\\s*([-+]?\\d*\\.?\\d+(?:[eE][-+]?\\d+)?)\\s*,\\s*([-+]?\\d*\\.?\\d+(?:[eE][-+]?\\d+)?)\\s*\\)")
  c(est = as.numeric(m[2]), lcl = as.numeric(m[3]), ucl = as.numeric(m[4]))
}

#--- (1) Build df of all methods ---------------
df_methods <- tibble(
  Method    = factor(c("Continuous-Unadjusted", "Continuous-Tian", "Continuous-Andersen"),
                     levels = c("Continuous-Unadjusted", "Continuous-Tian", "Continuous-Andersen")),
  Estimate  = c(cont_u_est, tian_est, andersen_est),
  Lower_CI  = c(cont_u_lcl, tian_lcl, andersen_lcl),
  Upper_CI  = c(cont_u_ucl, tian_ucl, andersen_ucl)
) %>%
  mutate(Estimate_CI = sprintf("%.2f (%.2f, %.2f)", Estimate, Lower_CI, Upper_CI))

#--- (2) Add RB-ANCOVA unadjusted & adjusted contrasts ---------------------
rb_tbl <- .get_tbl(results_1time)

rb_u  <- rb_tbl %>% filter(method == "unadjusted",
                           contrast == "Treated - Control") %>% slice(1)
rb_a  <- rb_tbl %>% filter(method == "covariate-adjusted",
                           contrast == "Treated - Control") %>% slice(1)

u_vals <- .parse_estCI(rb_u$est_CI)
a_vals <- .parse_estCI(rb_a$est_CI)

rb_df <- tibble(
  Method   = c("Unadjusted", "RB-ANCOVA"),
  Estimate = c(u_vals["est"], a_vals["est"]),
  Lower_CI = c(u_vals["lcl"], a_vals["lcl"]),
  Upper_CI = c(u_vals["ucl"], a_vals["ucl"])
)

#--- (3) Combine & order methods -------------------------------------------
plot_df <- bind_rows(df_methods, rb_df) %>%
  mutate(
    Estimate_CI = sprintf("%.2f (%.2f, %.2f)", Estimate, Lower_CI, Upper_CI),
    Method = factor(
      Method,
      levels = c("Continuous-Andersen", "Continuous-Tian", "RB-ANCOVA", "Continuous-Unadjusted", "Unadjusted")
    )
  )

# Collapse method names to two color groups
plot_df2 <- plot_df %>%
  mutate(
    Estimator = ifelse(grepl("unadjusted", Method, ignore.case = TRUE),
                       "Unadjusted", "Adjusted"),
    Estimator = factor(Estimator, levels = c("Unadjusted", "Adjusted")),
    Method = factor(Method, levels = levels(plot_df$Method)) # keep your chosen order
  )

pos <- position_dodge(width = 0.6)
```

```{r, echo=FALSE}
ggplot(plot_df2, aes(x = Method, y = Estimate, color = Estimator)) +
  geom_point(size = 4, shape = 16, position = pos) +
  geom_errorbar(aes(ymin = Lower_CI, ymax = Upper_CI),
                width = 0.2, position = pos, size = 1) +
  coord_flip() +
  scale_color_manual(values = c("Adjusted" = "blue", "Unadjusted" = "black")) +
  theme_minimal(base_size = 14) +
  labs(title = "Restricted Mean Survival Time Difference at 10 Years",
       x = NULL,
       y = "RMST Difference (Years)",
       color = "Estimator") +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
    axis.title = element_text(size = 14, face = "bold"),
    axis.text = element_text(size = 12),
    legend.position = "top",
    plot.margin = margin(5, 20, 5, 5)
  ) +
  #geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  # Label next to each point, matching line color
  geom_text(aes(label = Estimate_CI, color = Estimator),
            position = pos, vjust = 2, size = 4, hjust = 0.5, show.legend = FALSE) +
  scale_y_continuous(expand = expansion(mult = c(0.2, 0.1)))

```

***Figure 2.*** *Estimates and corresponding confidence intervals for difference in RMST of D-penicillamine vs. Placebo. Adjusted estimates adjust for histolic stage of disease, serum bilirubin (mg/dl), serum albumin (g/dl), and urine copper (ug/day). The Continuous-Adjusted results are shown from implementation of the continuous methodology, adjusting for covariates with Tian et al. (2018)'s ANCOVA-type method.*


## Multiple Timepoints

Due to the crossing of the KM curves for the placebo and PBC arms around 5 years, we additionally estimated the difference in RMST between the two groups for three time intervals of interest: 0 to 5 years, 0 to 10 years, and 5 to 10 years. This approach allowed us to assess hwo the treatment effect evolved over time. 

```{r}

results_2times <- rbancova_rmst(data_treated = pbc_filtered_trt, data_control=pbc_filtered_ctl, time_var="time_year", event_var="status_dichot", covariate_vars=c("stage","copper_imp", "bili", "albumin"), timepoints=c(5,10), num_intervals=10)

```

At 5 years, the unadjusted RMST difference (D-Penicillamine - placebo) was estimated to be 0.12 years with a 95\% confidence interval of [-0.20, 0.43]. This small difference suggests that the early phase of the trial showed minimal treatment effect, which is consistent with the observed survival probabilities during this period (***Figure 3***). Over 10 years, the unadjusted RMST difference was estimated to be -0.15 days with a 95\% confidence interval of [-0.95, 0.65], as described in the single timepoint analysis above. This estimate is in the opposite direction of the estimate from 0 to 5 years, with the switch happening between 5 and 10 years (***Figure 3***). The difference in RMST between 5 and 10 years was -0.27 years (95\% CI: [-0.83, 0.29]). All three confidence intervals include zero, indicating the possibility of no treatment effect from 0 to 5 years, 0 to 10 years, and from 5 to 10 years. 

To further refine these estimates, we employed RB-ANCOVA to adjust for covariates, as described in the single timepoint analysis above. After covariate adjustment, the RMST difference at 5 years was 0.01 years with a 95\% confidence interval of [-0.23, 0.24]. The adjusted RMST difference at 10 years was -0.44 days (95\% CI: [-1.05, 0.17]). The adjusted difference in RMST between 5 and 10 years was -0.44 years (95\% CI: [-0.91, 0.02]).These results indicate no significant differences between arms within any of the intervals (all 95% CIs include 0). Point estimates suggest minimal early separation at 5 years and a possible late disadvantage for the treated arm by 10 years, with the 5–10-year increment likewise negative. ***Figure 3*** illustrates these results, showing the unadjusted and adjusted RMST differences at both timepoints (5 years and 10 years) as well as the difference between them.

```{r, echo = FALSE, warning=FALSE, results=FALSE, message=FALSE}

res <- results_2times  # your object

# helper to pull numeric CI bounds from the "est_CI" string
get_bounds <- function(x) {
  m <- str_match(x, "\\(([-+]?\\d*\\.?\\d+(?:[eE][-+]?\\d+)?),\\s*([-+]?\\d*\\.?\\d+(?:[eE][-+]?\\d+)?)\\)")
  c(lower = as.numeric(m[,2]), upper = as.numeric(m[,3]))
}

# pull a single row (estimate + CI) from unadj/adj per-time or delta
grab <- function(which = c("unadj","adj"), tval = NULL, delta = FALSE) {
  which <- match.arg(which)
  src <- if (which == "unadj") res$unadjusted else res$adjusted
  stopifnot(!is.null(src))  # adjusted may be NULL if you ran without covariates
  
  df <- if (!delta) {
    src$per_time %>%
      filter(contrast == "Treated - Control",
             as.character(time) == as.character(tval)) %>%
      slice(1)
  } else {
    src$delta %>% slice(1)
  }
  b <- get_bounds(df$est_CI[1])
  tibble(Estimate = df$estimate[1],
         Lower_CI = b["lower"],
         Upper_CI = b["upper"])
}

# build long dataset in the order you described:
estimates_long <- bind_rows(
  mutate(grab("unadj", tval = 5,  delta = FALSE), Timepoint = "5 years",  Model = "Unadjusted"),
  mutate(grab("adj",   tval = 5,  delta = FALSE), Timepoint = "5 years",  Model = "Adjusted"),
  mutate(grab("unadj", tval = 10, delta = FALSE), Timepoint = "10 years", Model = "Unadjusted"),
  mutate(grab("adj",   tval = 10, delta = FALSE), Timepoint = "10 years", Model = "Adjusted"),
  mutate(grab("unadj", delta = TRUE),             Timepoint = "Difference (10y - 5y)", Model = "Unadjusted"),
  mutate(grab("adj",   delta = TRUE),             Timepoint = "Difference (10y - 5y)", Model = "Adjusted")
) %>%
  select(Timepoint, Model, Estimate, Lower_CI, Upper_CI) %>%
  mutate(
    Timepoint = factor(Timepoint,
                       levels = c("Difference (10y - 5y)", "10 years", "5 years")),
    Estimate_CI = sprintf("%.2f (%.2f, %.2f)", Estimate, Lower_CI, Upper_CI)
  )

estimates_long

# Plot
ggplot(estimates_long, aes(x = Timepoint, y = Estimate, color = Model)) +
  geom_point(size = 4, shape = 16, position = position_dodge(width = 0.6)) +  # Circles for both
  geom_errorbar(aes(ymin = Lower_CI, ymax = Upper_CI), 
                width = 0.2, position = position_dodge(width = 0.6), size = 1) +
  coord_flip() +
  scale_color_manual(values = c("blue", "black")) +
  theme_minimal(base_size = 14) +
  labs(title = "Restricted Mean Survival Time Difference Estimates at 5 and 10 Years",
       x = "Timepoint",
       y = "RMST Difference (Years)",
       color = "Estimator") +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
    axis.title = element_text(size = 14, face = "bold"),
    axis.text = element_text(size = 12),
    legend.position = "top",
    plot.margin = margin(5, 20, 5, 5)  # Adjust margins as needed
  ) +
  geom_vline(xintercept = 1.4, linetype = "dashed", size=1, color = "red")+
  #geom_hline(yintercept = 0, linetype = "dashed", color = "grey")+
  # Add Estimate (95% CI) below the bars
  geom_text(aes(label = Estimate_CI),
            position = position_dodge(width = 0.6),
            vjust = 2,  # Move text below the bar
            size = 4,
            hjust = 0.5, show.legend = F) +
  # Expand plot to ensure space for text below
  scale_y_continuous(expand = expansion(mult = c(0.2, 0.1)))

```


***Figure 3.*** *Estimates and corresponding confidence intervals for difference in RMST of PBC treatment vs. placebo at 5 years and 10 years and the difference between 5 and 10-year estimates. Adjusted estimates adjust for histologic stage of disease, copper (ug/day), serum bilirubin (mg/dl), and serum albumin (g/dl).*

## References

Andersen PK, Hansen MG, Klein JP. Regression analysis of restricted mean survival time based on pseudo-observations. Lifetime Data Analysis 2004;10(4):335–350.

Andersen PK, Klein JP, Rosthøj S. Generalised linear models for correlated pseudo-observations, with applications to multi-state models. Biometrika 2003;90:15–27.

Mayo Clinic Press Editors, Primary Biliary Cholangitis: Tests and Treatments You Can Expect; 2024. https://mcpress.mayoclinic.org/living-well/primary-biliary-cholangitis-tests-and-treatments-you-can-expect/, accessed: 2024-10-06.

Therneau TM. A Package for Survival Analysis in R; 2024, https://CRAN.R-project.org/package=survival, r package version 3.6-4

Therneau TM, Grambsch PM. Modeling Survival Data: Extending the Cox Model. New York: Springer; 2000

Tian L, Fu H, Ruberg SJ, Uno H, Wei LJ. Efficiency of Two Sample Tests via the Restricted Mean Survival Time for Analyzing Event Time Observations. Biometrics 2018;74(2):694–702. Epub 2017 Sep 12.


<!-- rmarkdown::render("vignettes/RMST-RBANCOVA.Rmd") -->