---
title: "RBANCOVA.RMST: RBANCOVA for RMST estimates with Categorized time-to-event data"
output:
  github_document:
    html_preview: true
vignette: >
  %\VignetteIndexEntry{RBANCOVA.RMST: RBANCOVA for RMST estimates with Categorized time-to-event data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
set.seed(2677)
```



```{r setup, include = FALSE}
# During development, you can uncomment the next line in an interactive session
# to use the functions in R/ without installing the package:
# devtools::load_all("..")
devtools::install()

library(RBANCOVA.RMST)
library(dplyr)

```

# Overview

This vignette shows how to implement Randmoziation-Based Covariance Analysis (RB_ANCOVA) for Restricted Mean Survival Time (RMST) contrasts using RBANCOVA.RMST. We’ll demonstrate via application to estimating the effect of D-Penicillamine for Primary Biliary Cholangitis (PBC).

## APPLICATION: D-PENICILLAMINE FOR PRIMARY BILIARY CHOLANGITIS
Primary biliary cholangitis (PBC) is a chronic autoimmune disease characterized by progressive inflammation and destruction of the small bile ducts within the liver, ultimately leading to cirrhosis (permanent scarring of liver tissues) and liver failure in advanced stages \citep{mayo2024}. The progression of PBC is inevitable, but it can be prolonged with treatment. Due to its clinical importance and the need for effective therapeutic options, PBC has been a focus of clinical trials for several decades. A dataset often used to study the natural history and treatment response in PBC patients originates from a Mayo Clinic trial conducted between 1974 and 1984.

The dataset includes 424 patients diagnosed with PBC who were referred to the Mayo Clinic between 1974 and 1984 \citep{survival-package, survival-book}. Of these, 312 patients were enrolled in a randomized, placebo-controlled trial of the drug D-penicillamine, while the remaining 112 cases did not participate in the clinical trial but consented to basic measurements and follow-up for survival. The dataset encompasses clinical variables, including age, sex, and a range of laboratory biomarkers such as serum bilirubin, serum albumin, and alkaline phosphatase levels, alongside clinical indicators like the presence of ascites, hepatomegaly, blood vessel malformations (spider angiomata), and histologic stage of disease at entry. Participants' survival outcomes are tracked until death, loss to follow-up (censored), liver transplantation (considered censored at time of transplant), or censoring at the study endpoint in July 1986. The PBC dataset can be accessed by installing the \emph{survival} package in R with dataset \emph{pbc} \citep{survival_package}.

```{r, include=TRUE}

#--------------------------------------------------------------------------------#
# Filter PBC Data
#--------------------------------------------------------------------------------#
library(survival)

#Filter to those enrolled in the randomized trial
pbc_filtered <- pbc %>% filter(!is.na(trt))

#Original data: 0/1/2 = censored/transplant/dead
##Edit so that transplant = censored at time of transplant [death = 1, 0 = censored/transplant]
pbc_filtered <- pbc_filtered %>%
  mutate(status_dichot = ifelse(status == 2, 1, 0))

#Create year variable and create arm variable [0=placebo, 1=D-penicillmain]
pbc_filtered <- pbc_filtered %>%
  mutate(time_year = time/365,
         arm = ifelse(trt == 2, 0, trt))

#Current RBANCOVA-RMST methodology assumes no missing data
#impute mean copper level within each group when missing. 
mean_copper_1 <- mean(pbc_filtered[!(is.na(pbc_filtered$copper)) & pbc_filtered$arm==1,]$copper)
mean_copper_0 <-  mean(pbc_filtered[!(is.na(pbc_filtered$copper)) & pbc_filtered$arm==0,]$copper)

pbc_filtered <- pbc_filtered %>%
  mutate(copper_imp = ifelse(is.na(copper), ifelse(arm==1, mean_copper_1, mean_copper_0), copper))

```


This analysis focuses on the 312 trial participants (158 in the D-penicillamine arm, 154 in the placebo arm). We implement randomization-based covariate adjustment for RMST estimation, as described in Section \ref{sec:ch4_methods}, to estimate treatment effects when adjusting for a selection of covariates. 


The methodology described in Section \ref{sec:ch4_rmst_methodology} was applied to the PBC dataset. The Kaplan-Meier curves illustrate the time to event for participants in the placebo and D-penicillamine groups over a 12.5-year follow-up period (Figure \ref{fig:pbc_km}). Over the first five years, there is minimal separation in survival probabilites between the two arms, with the D-penicillamine arm showing slightly higher survival probabilities compared to the placebo group. However, starting around five years, the Kaplan-Meier survival curves demonstrate that participants in the placebo arm generally had a higher probability of survival compared to those in the D-penicillamine group until approximately year nine or ten. %The separation between the curves becomes more noticeable around 5 years, indicating a potential survival disadvantage for the D-penicillamine group. 

```{r, echo = FALSE, warning=FALSE, results=FALSE, message=FALSE}

library(survminer)
fit_trt <- survfit(Surv(time_year, status_dichot) ~ trt,
                   data =pbc_filtered)
# Visualize with survminer
ggsurvplot(fit_trt, 
           data = pbc_filtered, 
           risk.table = TRUE, 
           censor.shape = NA,                 # Remove plus signs at censoring points
           legend.labs = c("D-penicillamine", "Placebo"),  # Rename the groups
           xlab = "Years",                     # Change x-axis label to 'Years'
           legend.title = "Trial Arm")    
```



<!-- \begin{figure} -->
<!-- \centering -->
<!-- \includegraphics[width=.8\linewidth]{Figures/pbc_survival_KM.png} -->
<!-- \caption{Kaplan-Meier (KM) estimate for time-to-death of each treatment group (D-penicillamine or placebo) for patients enrolled in the PBC Mayo Clinic trial. Time represents years from trial registration to death or last date known alive.}\label{fig:pbc_km} -->
<!-- \end{figure} -->

## Single Timepoint

To quantify the observed difference, we calculated the RMST for each group from the start of the trial until year ten using ten yearly intervals. Time was restricted to ten years of follow-up as many individuals at risk after year ten were censored before the end of follow-up. This yielded an RMST of 7.13 years (SE 0.28) for the D-penicillamine group and 7.28 years (SE 0.30) for the placebo group. %These results estimate that the average survival time during the first ten years of follow-up was 7.13 years in the D-penicillamine group and 7.28 years in the placebo group. The unadjusted difference in RMST (D-penicillamine - placebo) was estimated to be -0.15 years with a confidence interval of [-0.95, 0.65] (Figure \ref{fig:pbc_ci_1}). 

```{r}
pbc_filtered_trt <- pbc_filtered[pbc_filtered$arm==1,]
pbc_filtered_ctl <- pbc_filtered[pbc_filtered$arm==0,]

rbancova_rmst(data_treated = pbc_filtered_trt, data_control=pbc_filtered_ctl, time_var="time_year", event_var="status_dichot", covariate_vars=NULL, timepoints=10, num_intervals=10)

```

To further refine the estimated difference in RMST, we employed RB-ANCOVA to adjust for covariates selected via stepwise selection. The covariates retained for adjustment were histologic stage of disease, copper (ug/day), serum bilirubin (mg/dl), and serum albumin (g/dl), reflecting clinically relevant factors that might impact survival outcomes. Baseline balance between the trial arms was similar for most of these covariates (Table \ref{table:pbc_covs}): the average histoloic stage was slightly higher in the placebo group (3.09 vs. 2.97), while serum bilirubin levels were lower in the D-penicillamine group (2.87 vs. 3.65 mg/dl). Copper levels and serum albumin were comparable between groups, indicating minimal baseline differences. Assessing the extent of random imbalance between the two treatment arms, as described in Section \ref{sec:ch4_Step6_wls_main} gave no evidence of unexpected imbalance among the covariates (p-value = 0.23). 

<!-- \begin{table} -->
<!-- \caption{Mean (SE) of selected variables by treatment arm} -->

<!-- \centering -->
<!-- \begin{tabular}{lcccc} -->
<!-- \toprule -->
<!-- \textbf{Trial Arm} & \textbf{Histologic Stage} & \textbf{Copper} & \textbf{Serum Bilirubin} & \textbf{Serum Albumin}\\ -->
<!-- \midrule -->
<!-- Placebo & 3.09 (0.07) & 97.65 (6.46) & 3.65 (0.43) & 3.52 (0.03)\\ -->
<!-- D-penicillamine & 2.97 (0.07) & 97.64 (7.18) & 2.87 (0.29) & 3.52 (0.04)\\ -->
<!-- \hiderowcolors -->
<!-- \hline  % Please only put a hline at the end of the table -->
<!-- \end{tabular}\label{table:pbc_covs} -->
<!-- \end{table} -->


After adjusting for these covariates using RB-ANCOVA, the adjusted RMST difference was estimated to be -0.44 years with a confidence interval of [-1.05, 0.17].  
While the RB-ANCOVA results are consistent with the unadjusted results, the covariate adjustment improved the precision of the estimate. The CI half-width for the unadjusted estimate was 0.80 versus 0.61 for the adjusted estimate. 

```{r}

rbancova_rmst(data_treated = pbc_filtered_trt, data_control=pbc_filtered_ctl, time_var="time_year", event_var="status_dichot", covariate_vars=c("stage","copper_imp", "bili", "albumin"), timepoints=10, num_intervals=10)

```

Figure \ref{fig:pbc_ci_1} also shows the results of implementing the continuous methodology, unadjusted, as well as adjusted for covariates via \citet{Tian2018}'s ANCOVA-type method. Results when implementing the continuous methodology are consistent with those found with our methodology. Adjusting for covariates via RB-ANCOVA results in a similar estimate and confidence interval to that of \citet{Tian2018}'s method, but the RB-ANCOVA adjustment does not require the assumptions of \citet{Tian2018} to hold.


```{r, echo = FALSE}
#Figure for single time point
##Needs to get estimates from function
##Needs to get estimates from continuous function (Tian), maybe Andersen?
```

<!-- \begin{figure} -->
<!-- \centering -->
<!-- \includegraphics[width=0.8\linewidth]{Figures/estimates_CIs_pbc_4covs.png} -->
<!-- \caption{Estimates and corresponding confidence intervals for difference in RMST of D-penicillamine vs. Placebo. Adjusted estimates adjust for histolic stage of disease, serum bilirubin (mg/dl), serum albumin (g/dl), and urine copper (ug/day). The Continuous-Adjusted results are shown from implementation of the continuous methodology, adjusting for covariates with \cite{Tian2018}'s ANCOVA-type method.}\label{fig:pbc_ci_1} -->
<!-- \end{figure} -->

## Multiple Timepoints

<!-- rmarkdown::render("vignettes/RMST-RBANCOVA.Rmd") -->

```{r}

rbancova_rmst(data_treated = pbc_filtered_trt, data_control=pbc_filtered_ctl, time_var="time_year", event_var="status_dichot", covariate_vars=c("stage","copper_imp", "bili", "albumin"), timepoints=c(5,10), num_intervals=10)

```

## References

